---
title: "Count and Shift Tables"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Count_and_Shift_Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE, setup}
library(Tplyr)
library(dplyr, warn.conflicts = FALSE)
library(knitr)
library(kableExtra)
load("adsl.Rdata")
load("adae.Rdata")
load("adlb.Rdata")
```

# Frequency Tables

At the surface, counting sounds pretty simple, right? You just want to know how many occurrences of something there are. Well - unfortunately, it's not that easy. And in clinical reports, there's quite a bit of nuance that goes into the different types of frequency tables that need to be created. Fortunately, we've added a good bit of flexibility into both `group_count()` and `group_shift()` to help you get what you need when creating these reports, whether you're creating a demographics table, adverse events, of shift in lab results. 

### A simple example

Let's start with a basic example. This table demonstrates the distribution of subject disposition across treatment groups. Additionally, we're sorting by descending total occurrences using the `Total` group. 

```{r}
t <- tplyr_table(adsl, TRT01P, where = SAFFL == "Y") %>%
  add_total_group() %>%
  add_treat_grps(Treated = c("Xanomeline Low Dose", "Xanomeline High Dose")) %>%
  add_layer(
    group_count(DCDECOD) %>%
      set_order_count_method("bycount") %>%
      set_ordering_cols(Total)
  ) %>%
  build() %>%
  arrange(desc(ord_layer_1)) %>%
  select(starts_with("row"), var1_Placebo, `var1_Xanomeline Low Dose`,
         `var1_Xanomeline High Dose`, var1_Treated, var1_Total)

kable(t, "html") %>%
  column_spec(6, "3cm")
```

## Percentages

Counting is one thing - but a count by itself doesn't give much context. Often frequencies in clinical reports are paired with a percentage, which allows you to understand the proportion of events/subjects to some total - but that total can prove to be complicated, which is why we've included a few functions to aid in giving you the necessary denominator. 

  * `set_denoms_by()` function allows users calculate denominators using any variable within the by, cols, treatment, or target variables.  This defaults to all variables passed as treatment groups and column variables which causes the column percentages to sum to 100%. This is particularly useful for shift tables, which use a combination of row and column variables.
  * `set_denom_ignore()` defines values of the target variable to be ignored in the calculation of the denominator. In many tables percentages are based on non-missing counts. You can pass values to this function to have them be excluded from the calculation of the denominator. For example, if your data include NAs, you can use "NA" in `set_denom_ignore()` and those will be excluded from the denominator. 
  
Examples with percentages can be seen below.

## Distinct Versus Event Counts

Just as important as denominator flexibility is the determination of whether you should be using distinct or non-distinct counts - or some combination of both. Adverse event tables are a perfect example. Often, you're concerned about how many subjects had an adverse event in particular instead of just the number of occurrences of that adverse event. Similarly, the number occurrences of an event isn't necessarily relevant when compared to the total number of adverse events that occurred. For this reason, what you likely want to look at is instead the number of subjects who experienced an event compared to the total number of subjects in that treatment group. 

'Tplyr' allows you to focus on these distinct counts and distinct percents within some grouping variable, like subject. Additionally, you can mix and match with the non-distinct counts too. The `set_distinct_by` function sets the variables used to calculate the distinct occurrences of some value using the specified 'by' variables.

```{r}
t <- tplyr_table(adae, TRTA) %>%
  add_layer(
    group_count(AEDECOD) %>%
      set_distinct_by(USUBJID) %>%
      set_format_strings(f_str("xxx (xx.xx%) [xxx]", distinct, distinct_pct, n))
  ) %>%
  build() %>%
  head()

kable(t)
```

You may have seen tables before like the one about. This display shows the number of subjects who experienced an adverse event, the percentage of subjects within the given treatment group who experienced that event, and then the total number of occurrences of that event. Using `set_distinct_by()` triggered the derivation distinct_n and distinct_pct and the display of the values is then controlled by the `f_str()` call in `set_format_strings`.  

## Missing Counts

Missing counts are another tricky area for frequency tables. These values raise a number of questions. For example, do you want to format the missing counts the same way as the event counts? Do you want to present missing counts with percentages? Do missing counts belong in the denominator? 

The `set_missing_count()` function can take a new `f_str` object to set the display of missing values. Additionally, using the `string` parameter you can set the value chosen to be summarized as missing. By using a named character vector, you can control the row_label of what will be used for missing values.

In the example below 50 random values are removed and NA is specified as the missing string. The NAs are additionally ignored in denominators by using `set_denom_ignorE()`. Note that NA is submitted to these functions as a string even though the values are NAs in the data (i.e. not strings containing "NA"). During processing, the NAs are coerced into characters. 

```{r}
adae2 <- adae
adae2[sample(nrow(adae2), 50), "AESEV"] <- NA
t <- tplyr_table(adae2, TRTA) %>%
  add_layer(
    group_count(AESEV) %>%
      set_format_strings(f_str("xxx (xx.xx%)", n, pct)) %>%
      # AESEV of SEVERE is used to demonstrate missing
      set_missing_count(f_str("xxx", n), string = c(MISSING = "NA"))  %>%
      set_denom_ignore("NA")
    
  ) %>%
  build() %>%
  kable()

```

## Nested Count Summaries

Certain summary tables present counts within groups. One example could be in a disposition table where a disposition reason of "Other" summaries what those other reasons were. A very common example is an Adverse Event table that displays counts for body systems, and then the events within those body systems. This is again a nuanced situation - there are two variables being summarized: The body system counts, and the advert event counts.

One way to approach this would be creating two summaries. One summarizing the body system, and another summarizing the preferred terms by body system, and then merging the two together. But we don't want you to have to do that. Instead, we handle this complexity for you. This is done in `group_count()` by submitting two target variables with `dplyr::vars()`. The first variable should be your grouping variable that you want summarized, which we refer to as the "Outside" variable, and the second should have the narrower scope, which we call the "Inside" variable. 

The example below demonstrates how to do a nested summary. Look at the first row - here `row_label1` and `row_label2` are both "CARDIAC DISORDERS". This line is the summary for `AEBODSYS.` In the rows below that, `row_label1` continues on with the value "CARDIAC DISORDERS", but `row_label2` changes. These are the summaries for `AEDECOD`.

```{r}
tplyr_table(adae, TRTA) %>%
  add_layer(
    group_count(vars(AEBODSYS, AEDECOD))
  ) %>%
  build() %>%
  head() %>% 
  kable()
```

This accomplishes what we needed, but it's not exactly the presentation you might hope for. We have a solution for this as well.

```{r} 
tplyr_table(adae, TRTA) %>%
  add_layer(
    group_count(vars(AEBODSYS, AEDECOD)) %>% 
      set_nest_count(TRUE) %>% 
      set_indentation("--->")
  ) %>%
  build() %>%
  head() %>% 
  kable()
```

By using `set_nest_count()`, this triggers 'Tplyr' to drop row_label1, and indent all of the AEDECOD values within row_label2. The columns are renamed appropriately as well. The default indentation used with be 3 spaces, but as you can see here - you can set the indentation however you like. This let's you use tab strings for different language-specific output types, stick with spaces, indent wider or smaller - whatever you wish. All of the existing order variables remain, so this has no impact on your ability to sort the table. 

## Shifted Count Tables

Several tables include performing frequencies based on a shift of a baseline reference value. This is very common in laboratory tests where baseline references are compared with future references. These tables are a special case of count layers, but a method is provided that is intuitive and more efficient.

Constructing a shift layer is similar to constructing a count layer, however the reference values are passed using `dplyr::vars()` and named as 'row' and 'column.'

This example uses a randomly sampled adlb dataset which causes the table to appear sparse but the structure is demonstrated. The `LBNRIND` and `BNRIND` are updated to a factor to ensure they appear in the table output. If this wasn't done only 'NORMAL' and 'N' would appear in the dataset.
```{r}
adlb$LBNRIND <- factor(adlb$LBNRIND, c("NORMAL", "HIGH"))
adlb$BNRIND <- factor(adlb$BNRIND, c("N", "H"))
tplyr_table(adlb, TRTA, where = !is.na(LBNRIND) & VISIT == "WEEK 4") %>%
  add_layer(
    group_shift(vars(row = BNRIND, column = LBNRIND), by = vars(PARAM, VISIT))
  ) %>%
  build() %>%
  head() %>%
  kable()
```

A major part of the shift API is the control of the denominators used in the calculation of the percentages. In count layers most(but not all) tables have their percentages relative to the columns, each column sums to 100%. In shift layers, most percentages are relative to the "box" that is formed from the baseline and reference group for each treatment group. To support this, Tplyr allows you to pass the variables that the percentage denominators should be in reference too.

The `set_denoms_by()` functions can be passed unquoted variables names that were used in the table/layer construction. This will change the denominator of the percentages for the table.

```{r} 
suppressWarnings({
  tplyr_table(adlb, TRTA, where = !is.na(LBNRIND) & VISIT == "WEEK 4") %>%
  add_layer(
    group_shift(vars(row = BNRIND, column = LBNRIND), by = vars(PARAM, VISIT)) %>%
      set_format_strings(f_str("xx (xx.xx%)", n, pct)) %>%
      # This is what it normally defaults to, the 3x3 box formed by the target variables
      set_denoms_by(TRTA, PARAM, VISIT) 
    
  ) %>%
  build() %>%
  head() %>%
  kable()
})
```

In the example above, the denominators were based on the by and treatment variables, `TRTA`, `PARAM` and `VISIT`. In this example, the denominators are based on the 3 x 3 "box" formed by the baseline and reference variables.

Below, the denominators are changed to be based on the column instead of the 3 x 3 box. By passing the column variables, `TRTA` and `LBNRIND` the layer will use those denominator when determining the percentages.

```{r}
suppressWarnings({
  tplyr_table(adlb, TRTA, where = !is.na(LBNRIND) & VISIT == "WEEK 4") %>%
  add_layer(
    group_shift(vars(row = BNRIND, column = LBNRIND), by = vars(PARAM, VISIT)) %>%
      set_format_strings(f_str("xx (xx.xx%)", n, pct)) %>%
      set_denoms_by(TRTA, LBNRIND) # Use the column total as the denominator
  ) %>%
  build() %>%
  head() %>%
  kable()
})
```

In this example, the percentage denominators are calculated row-wise, each row percentage sums to 100%.

```{r}
suppressWarnings({
  tplyr_table(adlb, TRTA, where = !is.na(LBNRIND) & VISIT == "WEEK 4") %>%
  add_layer(
    group_shift(vars(row = BNRIND, column = LBNRIND), by = vars(PARAM, VISIT)) %>%
      set_format_strings(f_str("xx (xx.xx%)", n, pct)) %>%
      set_denoms_by(BNRIND, PARAM, VISIT) # Each row made by LBNRIND, TRTA
  ) %>%
  build() %>%
  head() %>%
  kable()
})
```

These are a few examples however you can pass any combination of variables used in the table or layer so any one of the 32 combinations of the five variables used in the example above could be used to calculate the denominators. 

# References
[Analysis and Displays Associated with Adverse Events](http://www.phusewiki.org/wiki/images/b/bf/Cs-whitepaper-adverseevents-v10-4442.pdf)

[Analyses and Displays Associated with Demographics, Disposition, and Medications](https://www.phuse.eu/documents//working-groups/deliverables/analyses-displays-associated-with-demographics-disposition-medications-in-phase-2-4-clinical-trials-version-20-02-mar-18-11808.pdf)

[Analyses and Displays Associated with Measures of Central Tendency](http://www.phusewiki.org/wiki/images/4/48/CSS_WhitePaper_CentralTendency_v1.0.pdf)
