---
title: "Creating Custom Tplyr Metadata"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{custom-metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(magrittr)
library(Tplyr)
library(knitr)
```

```{r data prep, echo=FALSE}
load("adas.Rdata")
load("adsl.Rdata")

t <- tplyr_table(adas, TRTP, where=EFFFL == "Y" & ITTFL == "Y" & PARAMCD == "ACTOT" & ANL01FL == "Y") %>% 
  set_pop_data(adsl) %>% 
  set_pop_treat_var(TRT01P) %>% 
  set_pop_where(EFFFL == "Y" & ITTFL == "Y") %>% 
  set_distinct_by(USUBJID) %>% 
  set_desc_layer_formats(
    'n' = f_str('xx', n),
    'Mean (SD)' = f_str('xx.x (xx.xx)', mean, sd),
    'Median (Range)' = f_str('xx.x (xxx;xx)', median, min, max)
  ) %>% 
  add_layer(
    group_desc(AVAL, where= AVISITN ==  0, by = "Baseline")
  ) %>% 
  add_layer(
    group_desc(AVAL, where= AVISITN == 24, by = "Week 24")
  ) %>% 
  add_layer(
    group_desc(CHG,  where= AVISITN == 24, by = "Change from Baseline")
  )

sum_data <- t %>% 
  build(metadata=TRUE) %>% 
  apply_row_masks(row_breaks = TRUE) %>% 
  select(row_id, starts_with('row_label'), 
         var1_Placebo, `var1_Xanomeline Low Dose`, `var1_Xanomeline High Dose`)

# I don't need the full model code for this example so just mock it up. 
# But if you want to see it, it's available here:
# https://github.com/RConsortium/submissions-pilot1/blob/694a207aca7e419513ffe16f6f5873526da1bdcb/R/eff_models.R#L17
model_portion <- tibble::tribble(
  ~"row_id",  ~"row_label1",                       ~"var1_Xanomeline Low Dose", ~"var1_Xanomeline High Dose",
  "x4_1",    "p-value(Dose Response) [1][2]",      "",                          "0.245",
  "x4_2",    "",                                   "",                          "",
  "x4_3",    "p-value(Xan - Placebo) [1][3]",	    "0.569",    	               "0.233",
  "x4_4",    "   Diff of LS Means (SE)",           "-0.5 (0.82)",               "-1.0 (0.84)",
  "x4_5",    "   95% CI",                          "(-2.1;1.1)",                "(-2.7;0.7)",
  "x4_6",    "",                                   "",                          "",
  "x4_7",    "p-value(Xan High - Xan Low) [1][3]", "",                          "0.520",
  "x4_8",    "   Diff of LS Means (SE)",           "",                          "-0.5 (0.84)",
  "x4_9",    "   95% CI",                          "",                          "(-2.2;1.1)"
)

full_data <- bind_rows(sum_data, model_portion) %>% 
  mutate(
    across(where(is.character), ~ replace_na(., ""))
  )



```

As covered in `vignette('metadata')`,**Tplyr** can produce metadata for any result that it calculates. But what about data that **Tplyr** can't produce, such as a efficacy results or some sort of custom analysis? You may still want that drill down capability either on your own or paired with an existing **Tplyr** table.

Take for instance Table 14-3.01 from the [CDISC Pilot](https://github.com/atorus-research/CDISC_pilot_replication). Skipping the actual construction of the table, here's the output data from **Tplyr** and some manual calculation:

```{r view data}
kable(full_data)
```

This is the primary efficacy table from the trial. The top portion of this table is fairly straightforward with **Tplyr** and can be done using descriptive statistic layers. Once you hit the p-values on the lower house, this becomes beyond Tplyr's remit. To produce the table, you can combine **Tplyr** output with a separate data frame analyzed and formatted yourself (but note you can still use some help from **Tplyr** tools like `apply_formats()`). 

But what about the metadata? How do you get the drill down capabilities for that lower half of the table? We've provided a couple additional tools in **Tplyr** to allow you to construct your own metadata and append existing metadata present in a **Tplyr** table.

## Build a `tplyr_meta` object

As covered in `vignette('metadata')`, a `tplyr_meta` object consists of two different fields: A list of variable names, and a list of filter conditions. You provide both of these fields as a list of quosures:

```{r tplyr_meta}
m <- tplyr_meta(
  names = quos(a, b, c),
  filters = quos(a==1, b==2, c==3)
)
m
```

The `tplyr_meta()` function can take these fields immediately upon creation. If you need to dynamically create a `tplyr_meta` object such as how **Tplyr** constructs the objects internally), the functions `add_variables()` and `add_filters()` are available to extend an existing `tplyr_meta` object:

```{r extending tplyr_meta}
m <- m %>% 
  add_variables(quos(x)) %>% 
  add_filters(quos(x == 'a'))

m
```

## Building your own metadata table

Now that we can create our own `tplyr_meta` objects, let's assemble the metadata for the bottom portion of Table 14-3.01:

```{r build efficacy metadata}
# Overall model subset of data
meta <- tplyr_meta(
  names = quos(TRTP, EFFFL, ITTFL, ANL01FL, SITEGR1, AVISIT, AVISITN, PARAMCD, AVAL, BASE, CHG),
  filters = quos(EFFFL == "Y", ITTFL == "Y", PARAMCD == "ACTOT", ANL01FL == "Y", AVISITN == 24)
)

# Xan High / Placebo contrast
meta_xhp <- meta %>% 
  add_filters(quos(TRTP %in% c("Xanomeline High Dose", "Placebo")))

# Xan Low / Placbo Contrast
meta_xlp <- meta %>% 
  add_filters(quos(TRTP %in% c("Xanomeline Low Dose", "Placebo")))

# Xan High / Xan Low Contrast
meta_xlh <- meta %>% 
  add_filters(quos(TRTP %in% c("Xanomeline High Dose", "Xanomeline Low Dose")))

eff_meta <- tibble::tribble(
  ~"row_id",  ~"row_label1",                       ~"var1_Xanomeline Low Dose", ~"var1_Xanomeline High Dose",
  "x4_1",    "p-value(Dose Response) [1][2]",      NULL,                        meta,
  "x4_3",    "p-value(Xan - Placebo) [1][3]",	     meta_xlp,    	              meta_xhp,
  "x4_4",    "   Diff of LS Means (SE)",           meta_xlp,                    meta_xhp,
  "x4_5",    "   95% CI",                          meta_xlp,                    meta_xhp,
  "x4_7",    "p-value(Xan High - Xan Low) [1][3]", NULL,                        meta_xlh,
  "x4_8",    "   Diff of LS Means (SE)",           NULL,                        meta_xlh,
  "x4_9",    "   95% CI",                          NULL,                        meta_xlh
)
```

Let's break down what happened here:

- First, we assemble the the overarching metadata object for the model. A lot of this metadata is shared across each of the different result cells for all of the efficacy data, so we can start by collecting this information into a `tplyr_meta` object.
- Next, we can use that starting point to build `tplyr_meta` objects for the other result cells. The model data contains contrasts of each of the different treatment group comparisons. By using `add_filters()`, we can create those additional three `tplyr_meta` objects using the starting point and attaching an additional filter condition.
- Lastly, to extend the metadata in the original `tplyr_table` object that created the summary portion of this table, we need a data frame. There's a lot of ways to do this, but I like the display and explicitness of `tibble::tribble()`.

When building a data frame for use with `tplyr_table` metadata, there are really only two rules:

- You need a column in the data frame called `row_id`
- The `row_id` values cannot be duplicates of any other value within the existing metadata.

The `row_id` values built by **Tplyr** will always follow the format "<first letter of layer type>n_n", where the first letter of the layer type will either be "c", "d", or "s". The next number is the layer number (i.e. the order in which the layer was inserted to the **Tplyr** table), and then finally the row of that layer within the output. For example, the third row of a count layer that was the second layer in the table would have a `row_id` of "c2_3". In this example, I chose "x4_n" as the format for the "x" to symbolize custom, and these data can be thought of as the fourth layer. That said, these values would typically be masked by the viewer of the table so they really just need to be unique - so you can choose whatever you want.

## Appending Existing **Tplyr** Metadata 

Now that we've created our custom extension of the **Tplyr** metadata, let's extend the existing data frame. To do this, **Tplyr** has the function `append_metadata()`:

```{r extending metadata}
t <- append_metadata(t, eff_meta)
```

Behind the scenes, this function simply binds the new metadata with the old in the proper section of the `tplyr_table` object. You can view the the `tplyr_table` metadata with the function `get_metadata()`:

```{r get_metadata}
get_metadata(t)
```

Finally, as with the automatically created metadata from Tplyr, we can query these result cells just the same:

```{r query custom metadata}
get_meta_subset(t, 'x4_1', "var1_Xanomeline High Dose") %>% 
  head() %>% 
  kable()
```

## Metadata Without Tplyr

You very well may have a scenario where you want to use these metadata functions outside of **Tplyr** in general. As such, there are S3 methods available to query metadata from a dataframe instead of a **Tplyr** table, and parameters to provide your own target data frame:

```{r metadata without Tplyr}
get_meta_subset(eff_meta, 'x4_1', "var1_Xanomeline High Dose", target=adas) %>% 
  head() %>% 
  kable()
```

As with the **Tplyr** metadata, the only strict criteria here is that your custom metadata have a `row_id` column.

## Tying it Together

The vignette wouldn't be complete without the final contextual example - so here we go. Ultimately these pieces an all fit together in the context of a Shiny application and give you the desired click-through experience.

```{r, out.width=850, out.extra='style="border: 1px solid #464646;" allowfullscreen="" allow="autoplay"', echo=FALSE}
knitr::include_app("http://michael-stackhouse.shinyapps.io/Tplyr-efficacy-shiny-demo", height = "900px")
```

_Source code available [here](https://github.com/atorus-research/Tplyr-efficacy-shiny-demo)_
